Hippocampal replay of experience at slow speeds

### 2021-04-22 Article.tex

In the stationary movement dynamic, the latent position does not change between time steps. The state transition matrix can thus be defined as an identity matrix, which predicts the next position will be the same as the last position (subject to the resolution of the position bin). In the continuous movement dynamic, the latent position is most likely to be "spatially close" to the position in the previous time step, so we use a Gaussian random walk state transition matrix. This means that, for a given latent position, the probability of moving to another position is modeled by a Gaussian centered at that position and "spatially close" is defined by the variance of the Gaussian. In our case, since replay has been shown to move at speeds much faster than the animal's speed \citep{DavidsonHippocampalReplayExtended2009, PfeifferAutoassociativedynamicsgeneration2015}, we set the variance of the Gaussian to 6.0 cm. This ensures that with a 2 ms time step, the latent position is 95\% likely to be within 4.90 cm of the previous latent position (or equivalently, this means that latent speeds from 0 to \texttildelow25 m/s are most likely).  Finally, in the fragmented movement dynamic, the latent position can move to any available position instantaneously. We model this using a uniform state transition matrix, which makes transitions to each position equally likely.

@@

An examination of the two summaries shown in \FIG{1}E reveals that that the model successfully captures the dynamics of the population spiking activity in \FIG{1}A. The stable firing of the one active neuron indicates a stationary representation, and accordingly, the probability of the stationary movement dynamic is high at the beginning of the simulated replay. A change in the data then drives a rapid transition to the continuous movement dynamic, reflecting the trajectory-like spatially sequential spiking from the population of cells. Subsequently, as the population activity becomes spatially incoherent, the fragmented movement dynamic dominates for the last 30 ms of the simulated event.

@@

@@

After testing our model on a single SWR event, we applied our clusterless decoding algorithm to hippocampal recordings from 10 rats performing the W-track spatial alternation task (\# tetrodes range: [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \citep{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020, CarrTransientSlowGamma2012}, position linearized to 1D). We confirmed that our decoding algorithm could closely track the position of the animal during behavior by comparing the most probable decoded position to the position of the animal (with the caveat that theta sweeps and replay sequences will represent positions other than where the animal is and our resolution is limited by the size of our 3 cm position bins). We found, similar to other studies, that the average median difference is small (4 cm median difference, 3-6 cm 95\% CI, for all time in sessions, 5 fold cross validation), but importantly, we achieved this using small time bins (2 ms) compared to the commonly used 250 ms time bin that most studies use for decoding during run and we were accurately able to decode immobility times as well, demonstrating the power of our algorithm.

@@

% Figure 5
\begin{figure}
\includegraphics[width=0.80\linewidth]{figures/Figure5/Figure5_v5}
\caption{
Prevalence of classifications. \textbf{(A)} UpSet plot \citep{LexUpSetVisualizationIntersecting2014}---which is similar to a Venn diagram with more than 3 sets---of the most common sets of classifications within each SWR. Each row represents a classification and each column represents a set of classifications, where filled-in black dots with an edge between the dots indicates that multiple classifications are present in the SWR (at temporally distinct times). The sets are ordered by how often they occur as indicated by the bar plot above each category. The total percentage of each classification is indicated by the rotated bar plot on the left. \textbf{(B)} Duration of each dynamic category within a SWR. The box shows the interquartile range (25-75\%) of the data and the whiskers are 1.5 times the interquartile range. Outlier points outside of this are labeled as diamonds. \textbf{(C)} Average distance between the latent position and the animal's actual position for each classification within the SWR. \textbf{(D)} Average speed of the classification within the SWR (excluding classifications with durations < 20 ms). Note that these speeds are calculated using the most probable position (MAP estimate) which can be noisy when the probability of position is flat or multimodal.\textbf{(E)} Kernel density estimate of the position of stationary trajectories on the W-track at least 30 cm away from the animal's position. The shaded region represents the density estimate while the black ticks represent the observed non-local stationary positions. \textbf{(F)} Average tetrode multiunit spike rates for each dynamic category within each SWR (excluding classifications < 20 ms).
}
\label{fig:5}

@@

Finally, we examined the overall levels of neuronal spiking across dynamics. We found that all the dynamics had similar average multiunit spiking rates (\FIG{5}F). This indicates that all dynamics, including stationary and stationary-continuous mixtures, were driven by sustained spiking information and could not be explained by the absence of spiking.

We performed several controls to see if our results were influenced by the firing of interneurons, which typically do not project to other brain areas and therefore cannot be read by a downstream region. First, we checked if requiring five or more tetrodes to have spikes during the SWR (up from requiring two tetrodes be active in the main results) would influence the proportions of dynamics, with the thought that a single tetrode with a high firing interneuron might result in stationary or stationary-continuous mixture dynamics during SWRs. This had little effect on the overall proportions, with many stationary and stationary-continuous mixtures still present (\FIGSUPP[5]{fig5sup2}A). Indeed, most of our events included spikes from upwards of 10 tetrodes (\FIGSUPP[5]{fig5sup3}B), including within each dynamic (\FIGSUPP[5]{fig5sup3}A). Next, we tried removing spikes that had spike widths of less than 0.3 ms because interneurons typically have narrow spike widths. Note that this is a rather imprecise threshold, because typically this is computed on the average waveform for a putative cell, not on a single spike, and combined with a requirement for the putative cell to have a high firing rate, as interneurons typically exhibit high firing rates. Nevertheless, even with this imprecise criteria, we still observed a high proportion of SWRs with stationary and stationary-continuous mixture dynamics (\FIGSUPP[5]{fig5sup2}B). We also looked at the 9 of 10 animals in our dataset that already had clustered cells, excluding putative interneurons based on spike width and firing rate and requiring that at least three putative pyramidal cells be active during the SWR. Although there were many fewer events to examine, we still observed many stationary and stationary-continuoux mixtures in SWRs compared to continuous sequences (\FIGSUPP[5]{fig5sup2}D). This provides further evidence that our results were not the result of the firing of interneurons.

@@

>>>>

In the stationary movement dynamic, the latent position does not change between time steps. The state transition matrix can thus be defined as an identity matrix, which predicts the next position will be the same as the last position (subject to the resolution of the position bin). In the continuous movement dynamic, the latent position is most likely to be "spatially close" to the position in the previous time step, so we use a Gaussian random walk state transition matrix. This means that, for a given latent position, the probability of moving to another position is modeled by a Gaussian centered at that position and "spatially close" is defined by the variance of the Gaussian. In our case, since replay has been shown to move at speeds much faster than the animal's speed \citep{DavidsonHippocampalReplayExtended2009, PfeifferAutoassociativedynamicsgeneration2015}, we set the variance of the Gaussian to 6.0 cm. This ensures that with a 2 ms time step, the latent position is 95\% likely to be within 4.90 cm of the previous latent position (or equivalently, this means that latent speeds from 0 to \texttildelow25 m/s are most likely). Last, in the fragmented movement dynamic, the latent position can move to any available position instantaneously. We model this using a uniform state transition matrix, which makes transitions to each position equally likely.

@@

An examination of the two summaries shown in \FIG{1}E reveals that that the model successfully captures the dynamics of the population spiking activity in \FIG{1}A. The stable firing of the one active neuron indicates a stationary representation, and accordingly, the probability of the stationary movement dynamic is high at the beginning of the simulated replay. A change in the data then drives a rapid transition to the continuous movement dynamic, reflecting the trajectory-like spatially sequential spiking from the population of cells. Subsequently, as the population activity becomes spatially incoherent, the fragmented movement dynamic dominates for the last 30 ms of the simulated event. This also shows that the fragmented dynamic gives us a way to test for spatially incoherent content directly for each time, rather than by the use of non-parametric shuffle, as is commonly used with the standard decoder.

@@

% Figure 2-Supplemental 3
\figsupp[Spatial coverage of the track for each animal.]{}{\includegraphics[width=0.5\linewidth]{figures/Figure2-supplemental3/Figure2_v1-supplemental3}}\label{figsupp:fig2sup3}

@@

After testing our model on a single SWR event, we applied our clusterless decoding algorithm to hippocampal recordings from 10 rats performing the W-track spatial alternation task (\# tetrodes range: [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \citep{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020, CarrTransientSlowGamma2012}, position linearized to 1D). We first confirmed that our cell population for each animal could closely track the position of the animal during behavior by comparing the most probable decoded position to the position of the animal (with the caveat that theta sequences and replay sequences will represent positions other than where the animal is and our resolution is limited by the size of our 3 cm position bins). We found, similar to other studies, that the average median difference is small (4 cm median difference, 3-6 cm 95\% CI, for all time in sessions, 5 fold cross validation). Importantly, we achieved this using small time bins (2 ms) compared to the commonly used 250 ms time bin that most studies use for decoding during run, demonstrating the power of our algorithm. We also verified that the population spiking for all positions on the track was consistent for each animal \FIGSUPP[2]{fig2sup3}.

@@

% Figure 5
\begin{figure}
\includegraphics[width=0.80\linewidth]{figures/Figure5/Figure5_v6}
\caption{
Prevalence of classifications. \textbf{(A)} UpSet plot \citep{LexUpSetVisualizationIntersecting2014}---which is similar to a Venn diagram with more than 3 sets---of the most common sets of classifications within each SWR. Each row represents a classification and each column represents a set of classifications, where filled-in black dots with an edge between the dots indicates that multiple classifications are present in the SWR (at temporally distinct times). The sets are ordered by how often they occur as indicated by the bar plot above each category. The total percentage of each classification is indicated by the rotated bar plot on the left. \textbf{(B)} Duration of each dynamic category within a SWR. The box shows the interquartile range (25-75\%) of the data and the whiskers are 1.5 times the interquartile range. Outlier points outside of this are labeled as diamonds. \textbf{(C)} Average distance between the latent position and the animal's actual position for each classification within the SWR. \textbf{(D)} Average speed of the classification within the SWR (excluding classifications with durations < 20 ms). Note that these speeds are calculated using the most probable position (MAP estimate) which can be noisy when the probability of position is flat or multimodal.\textbf{(E)} Kernel density estimate of the position of stationary trajectories on the W-track at least 30 cm away from the animal's position. The shaded region represents the density estimate while the black ticks represent the observed non-local stationary positions. \textbf{(F)} Average tetrode multiunit spike rates for each dynamic category within each SWR (excluding classifications < 20 ms).
}
\label{fig:5}

@@

We examined the overall levels of neuronal spiking across dynamics. We found that all the dynamics had similar average multiunit spiking rates (\FIG{5}F). This indicates that all dynamics, including stationary and stationary-continuous mixtures, were driven by sustained spiking information and could not be explained by the absence of spiking.

We performed several controls to see if our results were primarily influenced by the firing of interneurons, which typically do not project to other brain areas and may not be spatially specific in their firing (although note, their firing still may reflect relevant position information \ref{}). First, we checked if requiring five or more tetrodes to have spikes during the SWR (up from requiring two tetrodes be active) would influence the proportions of dynamics, with the thought that a single tetrode with a high firing interneuron might result in stationary or stationary-continuous mixture dynamics during SWRs. This had little effect on the overall proportions, with many stationary and stationary-continuous mixtures still present (\FIGSUPP[5]{fig5sup2}A). Indeed, most of our events included spikes from upwards of 10 tetrodes (\FIGSUPP[5]{fig5sup3}B), including within each dynamic (\FIGSUPP[5]{fig5sup3}A). Next, we tried removing spikes that had spike widths of less than 0.3 ms because interneurons typically have narrow spike widths. Note that this is a rather imprecise threshold, because typically this is computed on the average waveform for a putative cell, not on a single spike, and combined with a requirement for the putative cell to have a high firing rate, as interneurons typically exhibit high firing rates. Nevertheless, even with this imprecise criteria, we still observed a high proportion of SWRs with stationary and stationary-continuous mixture dynamics (\FIGSUPP[5]{fig5sup2}B). We also examined the 9 of 10 animals in our dataset that already had clustered cells, excluding putative interneurons based on spike width and firing rate and requiring that at least three putative pyramidal cells be active during the SWR. Although there were many fewer events to examine, we still observed many stationary and stationary-continuous mixtures in SWRs compared to continuous sequences (\FIGSUPP[5]{fig5sup2}D). Finally,  if interneurons were not spatially specific and driving the slower dynamics, we would expect to see the posterior be less spatially concentrated. Instead, we found that these dynamics tended to have narrow posteriors (as characterized by 95\% HPD region size in \FIG{4}E). This provides further evidence that our results were not the solely the result of the firing of interneurons.

@@

A third possible concern is that burst firing from a single cell is driving the stationary dynamics. There are several lines of evidence against this. First, as mentioned above, most of our SWRs had spikes from multiple tetrodes, including during stationary dynamics (\FIGSUPP[5]{fig5sup3}A), so it is unlikely that a single cell is driving these dynamics. Second, the stationary dynamics have a longer duration than one would expect from a burst alone (\FIG{5}B, 54 ms median duration, 38-74 ms quartiles vs. bursts which have a duration of 6-24 ms). Third, only ~50\% of the spikes had ISIs of less than 6 ms during the stationary, stationary-continuous, and continuous dynamics (\FIGSUPP[5]{fig5sup3}A), so if bursting occurs, it only occurs on half of the SWRs we have analyzed. It should be noted that this metric in particular should be interpreted with extreme caution, because with clusterless decoding, the ISI of the multiunit spikes is not the same as if we had an individual unit. The spikes could be from multiple neurons firing, which would not tell us if there was bursting or not. Nevertheless, even with these caveats, it seems unlikely that burst firing from cells is driving the stationary dynamics.
